Modular Type Constructors

### 2014-11-25 att-esop15/att-esop15.tex
Updated proposal

\renewcommand{\ttdefault}{txtt}
\usepackage{alltt}
\usepackage{listings}
\lstset{language=ML,
showstringspaces=false,
basicstyle=\ttfamily\footnotesize,
morekeywords={newcase,extends}}

@@

@@

\begin{abstract}
Abstraction providers sometimes need to extend a programming language with new base types and their corresponding operators. 
Ideally, such language fragments could be defined and reasoned about separately. Unfortunately, existing tools and techniques do not come equipped with strong modular reasoning principle -- fundamentally, %comparable to those available for library-based embeddings packaged using a modern module system. 
each combination of fragments becomes its own dialect of the language for which metatheoretic and compiler correctness results must be established monolithically. 
Recent work has started to address this problem, e.g. by showing how to modularly reason about desugarings atop a fixed type system. Our focus here is on safely composing type system fragments. We organize fragments around type constructors, as is usual practice when describing type systems, and sidestep the difficulties of composing abstract syntax (i.e. the expression problem)  by delegating control over a small, uniform abstract syntax in a type-directed manner to logic associated with a type constructor.
The paper is organized around a minimal calculus, @$\lambda$, that permits surprisingly practical examples. We establish several strong semantic guarantees, notably \emph{type safety}, \emph{stability of typing} under extension and \emph{conservativity}: 
that the \emph{type invariants} that a finite set of fragments maintain are conserved under  extension. 
This involves lifting typed compilation techniques into the semantics and enforcing an abstraction barrier around each fragment using a form of ``internal'' type abstraction. 
Conservativity then follows from classical parametricity results, so these \emph{modular type constructors} can be reasoned about like modules in an ML-like language. 
%\keywords{extensible languages; typed compilation; type-level computation; type abstraction}
\end{abstract}

\section{Introduction}
Typed programming languages are often described in \emph{fragments}, each defining contributions to a language's concrete syntax, abstract syntax, static semantics and dynamic semantics. 
In his textbook, Harper organizes fragments around type constructors, each introduced in a different chapter \cite{pfpl}. A language is then identified by a set of type constructors, e.g. $\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\keyw{1}}\,{\times}\,{+}\}$ is the language that builds in partial function types,  polymorphic types, recursive types, nullary and binary product types and binary sum types (its syntax is shown in Figure \ref{syntax-IL}, discussed further below).
It is left implied that the metatheory developed separately is conserved when fragments like these are composed to form a language.

Luckily, fragment composition is not an everyday programming task, because fragments like these are ``general purpose'' in that they make it possible to construct \emph{isomorphic embeddings} of many other fragments as ``libraries''. For example, lists can be placed in isomorphism with polymorphic recursive sum of products, $\iforall{\alpha}{\imu{t}{\iunit + (\alpha \times t))}}$. Languages providing datatypes in the style of ML  are perhaps most directly oriented around embeddings like this (preserving type disequality requires adding some form of generativity, as ML datatypes also expose). % Leaving the core language simple makes it easier to establish its metatheory and verify that tools, like compilers, are implemented correctly.% They also have encouraging connections to logic. %For example, we do not need to define the type constructor $\fvar{list}$ (indexed by a type) as a fragment (though it is possible to do so) because there is a user-defined datatype constructor, $\texttt{list}$, parameterized by a type, that is isomorphic. If we added the $\fvar{list}$ fragment to the language, it would be entirely redundant semantically: every well-typed term of a type constructed by $\fvar{list}$ corresponds to a well-typed term and a polymorphic recursive sum type such th%Although strings and numbers can be embedded as recursive sum types, it is recognized that this is impractical, so these are also usually included as primitives.
%Establishing an isomorphic embedding of a desirable fragment in terms of general-purpose fragments is not always possible, nor are such embeddings always practical.

Unfortunately, situations do sometimes arise where using these fragments to establish an isomorphic embedding that preserves a desirable fragment's  static and dynamic semantics (including bounds specified by its cost semantics) is not possible. 
Embeddings can also sometimes be unsatisfyingly \emph{complex}, as measured by the cost of the extralinguistic computations that are needed to map in and out of the embedding and, if these  must be performed mentally, considering  cognitive metrics like  error message comprehensibility. %We will discuss specific examples below.

When an embedding is too complex, abstraction providers have a few options, discussed in Sec. \ref{prior-work}. When an embedding is not possible, however, or when these options are  insufficient, providers are often compelled to introduce these fragments by extending an existing language, thereby forming a new \emph{dialect}. To save effort, they may do so by forking existing artifacts or leverage tools like compiler generators or language frameworks, also reviewed in Sec. \ref{prior-work}. %Indeed, the proliferation of language dialects constructed for this reason might be taken as an evidence that the core language is not as ``general'', when considered comprehensively, as might be hoped. 
Within the ML lineage, dialects that go  beyond $\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\keyw{1}}\,{\times}\,{+}\}$ abound:
%Reynolds, in a remark that recalls the ``Turing tarpit'' of Perlis \cite{Perl82a}, summarizes the issue \cite{Reynolds94anintroduction}: 
%\begin{quote}
%To say that any reasonable function can be expressed by some program is not to say that it can be expressed by the most reasonable program. It is clear that the language requires a novel programming style. Moreover, it is likely that certain important functions cannot be expressed by their most efficient algorithms.
%\end{quote}
%

\begin{enumerate}
\compresslist
\vspace{5px}
\item 
\textbf{General Purpose Fragments:} 
A number of variations on product types, for example, have been introduced in dialects: 
$n$-ary tuples, 
labeled tuples, 
records (identified up to reordering), 
structurally typed or row polymorphic records \cite{Cardelli:1984:SMI:1096.1098}, 
records with update and extension operators \cite{ocaml-manual}, 
mutable fields \cite{ocaml-manual}, 
%field delegation \cite{atlang-gpce14} \todo{gpce submission} 
and 
``methods'' (i.e. pure objects \cite{conf/oopsla/Aldrich13, TSLs}).\footnote{The Haskell wiki notes that ''No, extensible records are not implemented in GHC. The problem is that the record design space is large, and seems to lack local optima. And all reasonable variants break backward compatibility. As a result, nothing much happens.'' \cite{GHCFAQ}} 
 Sum types are also exposed in various ways: 
finite datatypes, 
open datatypes \cite{conf/ppdp/LohH06}, 
hierarchically open datatypes \cite{journals/toplas/MillsteinBC04}, 
polymorphic variants \cite{ocaml-manual} and 
ML-style exception types. Other generally useful fragments are also built in, e.g. \verb|sprintf| in the OCaml dialect statically distinguishes format strings from strings.

\item
\textbf{Specialized Fragments:} Fragments that track specialized static invariants to provide stronger correctness guarantees, manage unwieldy lower-level abstractions or control cost are also often introduced in dialects, e.g. for data parallelism  \cite{chakravarty2007data}, distributed programming \cite{Murphy:2007:TDP:1793574.1793585}, reactive programming \cite{mandel2005reactiveml}, authenticated data structures \cite{Miller:2014:ADS:2535838.2535851}, databases \cite{Ohori:2011:MSM:2034773.2034815} and units of measure \cite{conf/cefp/Kennedy09}.% All of these are implemented as dialects of existing languages, presumably because a strong encoding was not feasible.

@@

This dialect-oriented state of affairs is, we argue, unsatisfying: a programmer can choose either a dialect supporting a principled approach to distributed programming, or one that builds in support for statically reasoning about units of measure, but there may not be an available dialect supporting both. Combining dialects is non-trivial in general, as we will discuss. Using different dialects separately for different components of a program is also untenable: components written in different dialects cannot always interface safely with one another (i.e. an FFI, item 3  above, is needed).

These problems do not arise when a fragment can be exposed as an isomorphic embedding because modern \emph{module systems} can enforce abstraction barriers that ensure that the isomorphism need only be established in the ``closed world'' of the module. This is useful because it does not impose proof obligations on clients in the ``open world''.  %Mechanisms that can help decrease the complexity of an embedding without violating abstraction barriers are thus valuable, and we will lead into our work in Sec. \ref{prior-work} by summarizing them. 
For situations where an embedding is not evident, however, mechanisms are needed that make specifying, implementing and reasoning about direct fragment composition similarly modular. Such mechanisms could ultimately be integrated directly into a language, blurring the distinction between fragments and libraries and decreasing the need for new dialects. Importing fragments that introduce new syntax and semantics   would be as easy as importing a new module is today. In the limit, the community could rely on modularly mechanized metatheory and compiler correctness results, rather than requiring heroic efforts on the part of individual research groups as the situation exists today. %Recent work has shown progress on modularly introducing new concrete syntax, reviewed in Sec. \ref{desugaring}. Our focus is on the problem of introducing new semantics.

\paragraph{Contributions} In this paper, we take substantial steps towards this goal by  constructing a minimal but powerful core calculus, @$\lambda$ (the ``actively typed'' lambda calculus). %Despite its minimality, it can host a variety of practical semantic extensions like those described above, while maintaining strong metatheoretic guarantees and, crucially, providing modular reasoning properties. %a calculus introduced briefly in recent work on \emph{active type constructors}  \cite{atlang-gpce14}\todo{TR? Arxiv?}, reviewed in Sec. \ref{overview}. 
This calculus is structured in a manner similar to the Harper-Stone semantics for Standard ML \cite{Harper00atype-theoretic}, consisting of an \emph{external language} (EL) governed by a {typed translation semantics} targeting a fixed \emph{internal language} (IL). 
Rather than building in a monolithic set of type constructors, however, the typechecking judgement is indexed by a \emph{tycon context}. Each tycon defines the semantics of its associated operators via functions written in a \emph{static language} (SL). Types are values in the SL.%, which provides a form of type-level computation. %The authors demonstrate the expressive power of this technique primarily with an implementation and a number of examples of fragments as libraries, and their core calculus .% (unlike systems that treat the type system as a ``bag of rules'', where non-determinism can arise).

We introduce @$\lambda$ by example in Sec. \ref{overview}, demonstrating its surprising expressive power and detailing its semantics, then examine its key metatheoretic properties in Sec. \ref{calculus}, beginning with \emph{type safety} and \emph{unicity of typing} and touching on  \emph{decidability of typechecking}. We then consider properties that relate to composition of tycon definitions: \emph{hygiene}, \emph{stability of typing} and a key modularity result, which we refer to as  \emph{conservativity}: any invariants that can be established about all values of a type under \emph{some} tycon context (i.e. in some  ``closed world'') are conserved in any further extended tycon context (i.e. in the ``open world'').

We combine several interesting type theoretic techniques, applying them to novel ends: 1)  a bidirectional type system permits flexible reuse of a fixed syntax; 2) the SL serves both as an extension language and as the type-level language; we give it its own statics (i.e. a \emph{kind system}); 3) we use a typed intermediate language and leverage corresponding \emph{typed compilation} techniques, here lifted into the semantics of the EL; 4) we leverage internal type abstraction implicitly as an effect during normalization of the SL to enforce abstraction barriers between type constructors. 
As a result, conservativity follows from the same elegant parametricity results that underly  abstraction theorems for module systems. 
Like modules, reasoning about these \emph{modular type constructors} does not require  mechanized specifications or proofs: correctness issues in the type constructor logic necessarily causes typechecking to fail, so even extensions that are not proven correct can be distributed and ``tested'' in the wild without compromising the integrity of an entire program (at worst, only values of types constructed by the tycon being tested may exhibit undesirable properties). 
% But mechanized proofs can be provided to verify that such failures (which do not indicate mistakes in client code but are more analagous to ``translation validation failures'' arising from the compiler \cite{Pnueli-Siegel-Singerman98}, will not occur during typechecking), 
% and that a type constructor adequately implements a separately specified fragment. We take the first steps towards such a \emph{modularly mechanized metatheory} by briefly discussing how to construct contextual embeddings of our semantics into Coq (Sec. \ref{coq}). 
%Though surprisingly expressive, we discuss some limitations of the core calculus, and suggest variants  that increase its expressive power while maintaining these key guarantees (Sec. \ref{variants}). 
We conclude by discussing some limitations, proposing some richer variants of the calculus and discussing related work (Sec. \ref{prior-work}). %Note that readers unfamiliar with prior approaches may wish to read this section first. %so that even fragments like those enumerated above can be defined as safely composable libraries, rather than in dialects that must be considered monolithically. %In the limit, building safe FFIs between every pair languages in a project is more effort than attempting to combine the languages. 
%Libraries are preferable to dialects, so there has been considerable interest in mechanisms that can make it possible to implement and reason about fragments like those above orthogonally and compose them automatically.% Mechanisms that guarantee composition will \emph{always} be safe can be integrated directly into a language as a general-purpose abstraction mechanism.  %that can fully automate the process of combining dialects defined atop a common framework (\emph{language-external mechanisms}).

% CANT GUARANTEE THAT SPECIFICATIONS ARE ACTUALLY DECIDABLE 
\bibliographystyle{abbrv}
\bibliography{../research}
\appendix
\section{Appendix}

\end{document}

>>>>

\usepackage{times}
\renewcommand{\ttdefault}{txtt}
\usepackage{alltt}
\usepackage{listings}
\lstset{language=ML,
showstringspaces=false,
basicstyle=\ttfamily\footnotesize,
morekeywords={newcase,extends}}

@@

\setlength{\abovecaptionskip}{5px}
\setlength{\belowcaptionskip}{5px}

\newcommand{\moutput}{^{\color{gray}+}}
\newcommand{\rulename}[1]{(#1)}
\def \TirNameStyle #1{\small\rulename{#1}}
\renewcommand{\MathparLineskip}{\lineskiplimit=.6\baselineskip\lineskip=.6\baselineskip plus .2\baselineskip}

@@

\begin{abstract}
Researchers commonly describe typed programming languages in fragments or as simple calculi, leaving to language designers the task of composing these to form complete languages.  %comparable to those available for library-based embeddings packaged using a modern module system. 
Metatheoretic results must be established monolithically for each such language, guided only informally by the results derived for simpler calculi.
We argue that, as the language design space grows, mechanisms for defining and composing typed language fragments that provide strong modular reasoning principles are needed.%, so that important metatheoretic results need only be established locally and, ultimately, so that language extensions can be treated like libraries. 
%,  Sometimes this fails because certain language features do not ``play well'' together, but precisely characterizing which is elusive given these practices.

%Recent work has started to address this problem, e.g. by showing how to modularly reason about desugarings atop a fixed type system. Our focus here is on safely composing type system fragments. We organize fragments around type constructors, as is usual practice when describing type systems, and sidestep the difficulties of composing abstract syntax (i.e. the expression problem)  by delegating control over a small, uniform abstract syntax in a type-directed manner to logic associated with a type constructor.

In this paper, we begin from first principles with a minimal calculus, @$\lambda$, specified as a bidirectionally typed translation semantics: external terms are simultaneously typechecked and given a translation to a fixed typed internal language. The external language is made extensible by indexing the typing judgement by a \emph{tycon context}. Each tycon defines the semantics of its associated operators using a fixed static language where types are values. This language is constructed to ensure that several strong semantic guarantees follow from reasoning that can be performed in the ``closed world'' of a fixed tycon context, notably \emph{type safety} and \emph{conservativity}: that the \emph{type invariants} maintained under a minimal tycon context will be conserved under any further extended context. 
Violations are caught during typechecking by lifting typed compilation techniques into the semantics and enforcing an abstraction barrier around each tycon using type abstraction, so these \emph{modular type constructors} can be reasoned about modularly, i.e. like modules in an ML-like language. 
%\keywords{extensible languages; typed compilation; type-level computation; type abstraction}
\end{abstract}

\section{Introduction}
Typed programming languages are often described in \emph{fragments}, each defining distinct contributions to a language's concrete and abstract syntax, static semantics and dynamic semantics. 
For example, in his textbook, Harper organizes fragments around type constructors, describing each in a different chapter \cite{pfpl}. Complete languages are then identified by a set of type constructors, e.g. $\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\keyw{1}}\,{\times}\,{+}\}$ is the language that builds in partial function types,  polymorphic types, recursive types, nullary and binary product types and binary sum types (its syntax is shown in Figure \ref{syntax-IL}, discussed further below).
Another common pattern is for a researcher to describe a new fragment by defining a complete calculus, perhaps including a ``catch-all'' constant and base type to stand for all other terms and types that may be of interest.

Metatheoretic reasoning techniques for programming languages generally assume that a complete specification of a language is being considered, however. To use these techniques, every \emph{combination} of fragments must be treated as its own monolithic language for which metatheorems must be established anew, guided only informally by results derived for the smaller systems from which they are, notionally, composed.

%In both cases, it is left as an extrinsic concern to ensure that the metatheory developed separately is actually conserved when fragments defined in these ways are composed to form a complete language.

Luckily, fragment composition is not an everyday programming task because fragments like these are ``general purpose'' in that they make it possible to construct \emph{isomorphic embeddings} of many other fragments as ``libraries''. For example, lists need not be built in ``primitively'' because they can be placed in isomorphism with the polymorphic recursive sum of products $\iforall{\alpha}{\imu{t}{\iunit + (\alpha \times t))}}$. Languages providing datatypes al\'a ML  are perhaps most directly oriented around such embeddings. % Leaving the core language simple makes it easier to establish its metatheory and verify that tools, like compilers, are implemented correctly.% They also have encouraging connections to logic. %For example, we do not need to define the type constructor $\fvar{list}$ (indexed by a type) as a fragment (though it is possible to do so) because there is a user-defined datatype constructor, $\texttt{list}$, parameterized by a type, that is isomorphic. If we added the $\fvar{list}$ fragment to the language, it would be entirely redundant semantically: every well-typed term of a type constructed by $\fvar{list}$ corresponds to a well-typed term and a polymorphic recursive sum type such th%Although strings and numbers can be embedded as recursive sum types, it is recognized that this is impractical, so these are also usually included as primitives.
%Establishing an isomorphic embedding of a desirable fragment in terms of general-purpose fragments is not always possible, nor are such embeddings always practical.

Although these fragments certainly appear to occupy a ``sweet spot'' in the design space, ``general purpose'' does not mean ``all-purpose'' and situations do continue to arise where using these fragments to establish an isomorphic embedding that preserves a desirable fragment's  static and dynamic semantics (including bounds specified by a cost semantics) is not possible. 
Embeddings can also sometimes be unsatisfyingly \emph{complex}, as measured by the cost of the extralinguistic computations that are needed to map in and out of the embedding and, if these  must be performed mentally by programmers, considering  cognitive factors. %We will discuss specific examples below.
%When an embedding is too complex, abstraction providers have a few options, discussed in Sec. \ref{prior-work}. When an embedding is not possible, however, or when these options are  insufficient, 
%providers are often compelled to introduce these fragments by extending an existing language, thereby forming a new \emph{dialect}. To save effort, they may do so by forking existing artifacts or leverage tools like compiler generators or language frameworks, also reviewed in Sec. \ref{prior-work}. %Indeed, the proliferation of language dialects constructed for this reason might be taken as an evidence that the core language is not as ``general'', when considered comprehensively, as might be hoped. 
Each time a researcher seeks to address one of these problems by designing a new language construct, a new \emph{dialect} of the language is born. Within the ML lineage, dialects that go  beyond core ML abound:
%Reynolds, in a remark that recalls the ``Turing tarpit'' of Perlis \cite{Perl82a}, summarizes the issue \cite{Reynolds94anintroduction}: 
%\begin{quote}
%To say that any reasonable function can be expressed by some program is not to say that it can be expressed by the most reasonable program. It is clear that the language requires a novel programming style. Moreover, it is likely that certain important functions cannot be expressed by their most efficient algorithms.
%\end{quote}
%

\begin{enumerate}
%\compresslist
\item 
\textbf{General Purpose Fragments:} 
A number of variations on product types, for example, have been introduced in dialects: 
$n$-ary tuples, 
labeled tuples, 
records (identified up to reordering), 
structurally typed or row polymorphic records \cite{Cardelli:1984:SMI:1096.1098}, 
records with update and extension operators \cite{ocaml-manual}, 
mutable fields \cite{ocaml-manual}, 
%field delegation \cite{atlang-gpce14} \todo{gpce submission} 
and 
``methods'' (i.e. pure objects \cite{TSLs}).\footnote{The Haskell wiki notes that ''No, extensible records are not implemented in GHC. The problem is that the record design space is large, and seems to lack local optima. [...] As a result, nothing much happens.'' \cite{GHCFAQ}} 
 Sum-like types are also exposed in various ways: 
finite datatypes, 
open datatypes \cite{conf/ppdp/LohH06}, 
hierarchically open datatypes \cite{journals/toplas/MillsteinBC04}, 
polymorphic variants \cite{ocaml-manual} and 
ML-style exception types. Other generally useful fragments are also built in, e.g. \verb|sprintf| in the OCaml dialect statically distinguishes format strings from strings.

\item
\textbf{Specialized Fragments:} Fragments that track specialized static invariants to provide stronger correctness guarantees, manage unwieldy lower-level abstractions or control cost are also often introduced in dialects, e.g. for data parallelism  \cite{chakravarty2007data}, distributed programming \cite{Murphy:2007:TDP:1793574.1793585}, reactive programming \cite{mandel2005reactiveml}, authenticated data structures \cite{Miller:2014:ADS:2535838.2535851}, databases \cite{Ohori:2011:MSM:2034773.2034815},  units of measure \cite{conf/cefp/Kennedy09} and string sanitation \cite{sanitation-psp14}.% All of these are implemented as dialects of existing languages, presumably because a strong encoding was not feasible.

@@

This \emph{dialect-oriented} state of affairs is unsatisfying for language  designers, fragment providers and programmers alike. Language designers are burdened with needing to understand the complete metatheory of every fragment they add to their language to make sure it is not incompatible with existing fragments, so languages change rarely. This decreases the impact of many potentially useful innovations designed by fragment providers (e.g. many academic researchers) by making them unavailable to programmers. At best, programmers can choose from either a dialect supporting, for example, a principled approach to distributed programming, or one that builds in support for statically reasoning about units of measure. There may not be an available dialect supporting both. Using different dialects separately for different components of a program is untenable: components written in different dialects cannot always interface safely (an FFI, above, is needed).

These problems do not arise when a fragment can be expressed as an isomorphic embedding (i.e. as a library) because modern \emph{module systems} enforce abstraction barriers that ensure that the isomorphism need only be established in the ``closed world'' of the module. This is useful because it does not impose proof obligations on clients in the ``open world''. For example, a module defining the semantics of sets in ML can hold the representation of sets abstract, ensuring that any invariants maintained by the functions in the module (e.g. uniqueness) will hold even when other modules are in use. %Mechanisms that can help decrease the complexity of an embedding without violating abstraction barriers are thus valuable, and we will lead into our work in Sec. \ref{prior-work} by summarizing them.

Because embedding of this form are not always possible, as in the examples enumerated above, so mechanisms are needed that make it possible to modularly define new fragments that more directly influence the static and dynamic semantics of a language. Such mechanisms could ultimately be integrated directly into a language, blurring the distinction between fragments and libraries and decreasing the need for new dialects. Importing fragments that introduce new syntax and semantics   would be as safe and easy as importing a new module is today. %In the limit, the community could rely on modularly mechanized metatheory and compiler correctness results.% rather than requiring heroic efforts from individual research groups that consider an entire language at once. %Recent work has shown progress on modularly introducing new concrete syntax, reviewed in Sec. \ref{desugaring}. Our focus is on the problem of introducing new semantics.
\begin{figure}[t]
%\small
\input{syntax-table-IL}
\caption{Syntax of {$\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\iunit}\,{\times}\,{+}\}$}, our internal language (IL).}
\label{syntax-IL}
%\vspace{-10px}
\end{figure}

\vspace{-10px}
\subsubsection{Contributions} In this paper, we take substantial steps towards this goal by  constructing a minimal but powerful core calculus, @$\lambda$ (the ``actively typed'' lambda calculus). %Despite its minimality, it can host a variety of practical semantic extensions like those described above, while maintaining strong metatheoretic guarantees and, crucially, providing modular reasoning properties. %a calculus introduced briefly in recent work on \emph{active type constructors}  \cite{atlang-gpce14}\todo{TR? Arxiv?}, reviewed in Sec. \ref{overview}. 
This calculus is structured in a manner similar to the Harper-Stone semantics for Standard ML \cite{Harper00atype-theoretic}, consisting of an \emph{external language} (EL) governed by a {(bidirectionally) typed translation semantics} targeting a fixed \emph{internal language} (IL). 
Rather than building in a monolithic set of external type constructors, however, the typechecking judgement is indexed by a \emph{tycon context}. Each tycon defines the semantics of its associated operators via functions written in a \emph{static language} (SL). Types are values in the SL.%, which provides a form of type-level computation. %The authors demonstrate the expressive power of this technique primarily with an implementation and a number of examples of fragments as libraries, and their core calculus .% (unlike systems that treat the type system as a ``bag of rules'', where non-determinism can arise).

The bulk of the paper, in Sec. \ref{atlam}, introduces @$\lambda$ by building up two examples, one defining labeled product types with a functional update operator,  and the other regular string types, based on a recent ``core calculus'' style specification \cite{sanitation-psp14}.  We then examine the key metatheoretic properties of the calculus in Sec. \ref{metatheory}, beginning with \emph{type safety} and \emph{unicity of typing}, touching on  \emph{decidability of typechecking} and finally considering properties that relate to composition of tycon definitions: \emph{hygiene}, \emph{stability of typing} and a key modularity result, which we refer to as  \emph{conservativity}: any invariants that can be established about all values of a type under \emph{some} tycon context (i.e. in some  ``closed world'') are conserved in any further extended tycon context (i.e. in the ``open world''). Interestingly, the approach we take to guarantee conservativity relies on type abstraction, in this case in the internal language. As a result, we are able to borrow the same results that underly modular reasoning in languages like ML to reason about extensions to the semantics itself. Finally, we describe  related work in Sec. \ref{prior-work}.

\section{@$\lambda$}\label{atlam}
\begin{figure}[t]
%\small
\hspace{-2px}\input{syntax-table-EL}
\caption{Syntax of the external language (EL). We write $x$ to range over variables and $\opname{op}$ over operator names. The form $\eother{\iota}$ is a technical device (and would have no concrete syntax).}\label{syntax-EL}
\end{figure}

We will begin by giving an overview of the basic organization of @$\lambda$ and introduce the main judgements in Section \ref{overview}, discuss how types are constructed and tycons are defined in Section \ref{types}, then describe how the semantics of external terms are controlled by tycons in Section \ref{external-terms}.

\subsection{Overview}\label{overview}
\vspace{-5px}
\subsubsection{External Language}
Programs in @$\lambda$ are written as \emph{external terms}, $e$. The syntax of external terms is shown in Figure \ref{syntax-EL}. %We will describe useful syntactic desugarings atop this syntax as we go on and review recent techniques that permit modularly introducing  desugarings like these in Sec. \ref{desugaring}. Our main focus here is on semantic extensions. 
The static and dynamic semantics of external terms are specified simultaneously as a \emph{bidirectionally typed translation semantics}. The two central judgements in this paper take the form:  \[\esynX{e}{\st\moutput}{\iota\moutput} ~~~~~\text{and}~~~~~ \eanaX{e}{\st}{\iota\moutput}\]
\noindent

These are pronounced ``$e$ (synthesizes / analyzes against) type $\sigma$ and has  translation $\iota$ under typing context $\Upsilon$ and tycon context $\Phi$''. Our specifications in this paper are intended to be algorithmic: we indicate ``outputs'' when introducing judgement forms by \emph{mode annotations}, $\moutput$; these are not part of the judgement's syntax.

Bidirectional typechecking, also sometimes called \emph{local type inference} \cite{Pierce:2000:LTI:345099.345100}, is increasingly being used in modern languages (e.g. Scala) because it eliminates the need for type annotations in many circumstances while remaining decidable in more situations than Hindley-Milner style inference and providing what are widely perceived to be higher quality error messages, due to the locality of reasoning. It will also give us a clean way to reuse a fixed set of  introductory forms, discussed in Sec. \ref{external-terms}.

%We will return to how we control the semantics of the single introductory form in the abstract syntax, $\eintro{\st}{\es}$, on the basis of the type it is being analyzed against, sidestepping issues related to the \emph{expression problem} as a result \todo{cite other paper}\cite{wadler1998expression}. The form $\eother{\iota}$ is a technical device used to quantify over all terms that may arise in a ``future'' tycon context, serving an analagous role to the  ``catchall'' base constants commonly found in minimal calculi, and we will return to its semantics later as well.%We will see how we leverage this bidirectionality when we discuss literal forms below. % The judgement is only well-defined for \emph{valid} tycon contexts, written judgementally as $\vdash \Phi$ and \emph{valid} typing contexts, $\vdash_\Phi \Upsilon$, which we also specify below.
%  \begin{figure*}[t!]
% \small
% \hspace{-5px}\input{example-table}
% %\\~\\
% %$\begin{array}{rcl}
% %e\cdot\conclbl{lbl}(\conclbl{lbl}_1=e_1, \cdots, \conclbl{lbl}_n=e_n)  :=  \etarg{(\conclbl{lbl}; \svar{list}n[\klbl]~\conclbl{lbl}_1~\cdots~\conclbl{lbl}_n)}{e}{e_1; \cdots; e_n}
% %\end{array}$
% \caption{An example external term, $e_{ex}$, in concrete syntax (left), desugared to abstract syntax (right), with static terms shown in green in examples (only). The notation ${\small \scolor{\desugar{...}}}$ stands for an embedding of the indicated ${\small \scolor{\conclbl{label}}}$, ${\small \scolor{\concrx{regular expression}}}$, ${\small \scolor{\concstr{string}}}$ or numeral into the SL, with derived kinds  $\klbl$, $\krx$,  $\kstr$ and $\knat$, not shown. The signatures of helper functions, e.g. ${\small \scolor{\svar{nil}}}$ and ${\small \scolor{\svar{list}n}}$, are shown in Figure \ref{helper-sigs}.}
% \label{example}
% \end{figure*}

\vspace{-10px}
\subsubsection{Internal Language} Internal terms, $\iota$, together with internal types, $\tau$, form the \emph{typed internal language}. @$\lambda$ relies on a typed internal language supporting type abstraction (i.e. universal quantification over types). In this paper, we use {$\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\iunit}\,{\times}\,{+}\}$}, the syntax for which is shown in Figure \ref{syntax-IL}. This is representative of a  standard intermediate language as might be found for a language like ML. The IL has a fixed semantics, so different choices of IL represent different dialects of @$\lambda$.

We assume the internal statics are specified in the standard way by judgements for internal type formation {$\iDelta \vdash \tau$}, typing context formation { $\iDelta \vdash \iGamma$} and type assignment {$\iDelta~\iGamma \vdash \iota : \tau\moutput$}. The typing and type formation contexts obey standard structural properties (i.e. weakening, exchange, contraction). Throughout the paper, we will omit leading $\emptyset$ (used as the base case for finite mappings) and $\cdot$ (used as the base case for finite sequences) in examples (e.g. writing $\Gamma_\text{1} := x : \tau$ rather than $\Gamma_\text{1} := \emptyset, x : \tau$).

The internal dynamics are also a standard structural operational semantics with stepping judgement {\small $\iota \mapsto \iota\moutput$} and value judgement {$\iota~\mathtt{val}$}. The judgement $\iota \mapsto^{*} \iota\moutput$ is the reflexive, transitive closure of the stepping judgement. The semantics of the IL can be found in any standard textbook covering typed lambda calculi (we closely follow \cite{pfpl}), so we assume familiarity and omit the details.

% We also here define a syntax for simultaneous substitutions, of terms for variables, $\gamma$, and of types for polymorphic type variables, $\delta$, and assume standard judgements ensuring that these are valid with respect to a valid context, $\Delta \vdash \gamma : \Gamma$ and $\vdash \delta : \Delta$. We apply these substitutions to terms, types and typing contexts using the syntax $[\gamma]\iota$, $[\delta]\iota$, $[\delta]\tau$ and $[\delta]\Gamma$ (in some prior work, application of a substitution like this is written using a hatted form, e.g. $\hat\gamma(\iota)$; we intend the same semantics but use a notation more consistent with standard substitution, e.g. $[\iota/x]\iota'$). We will omit  leading $\emptyset$ and $\cdot$ in examples; in specifications, the former is used for  metatheoretic finite mappings, the latter for metatheoretic ordered lists.

\vspace{-10px}
\subsubsection{Static Language}
\begin{figure}[t]
\input{syntax-table-SL}
\caption{Syntax of the static language (SL).}
\vspace{-10px}
\label{syntax-SL}
\end{figure}
The workhorse of our calculus is the \emph{static language}, which itself forms a typed lambda calculus where 
%External terms are classified by (external) \emph{types}. 
\emph{kinds}, $\kappa$, classify \emph{static terms}, $\sigma$.  The syntax of the SL is given in Figure \ref{syntax-SL}. We note that the core of the SL is a total functional programming language based  on several standard fragments: total functions, quantification over kinds, inductive kinds (constrained by a positivity condition to prevent non-termination), and products and sums. These all also closely follow \cite{pfpl}, so we again omit some details. Only three new kinds are introduced: $\kty$, $\kity$ and $\kitm$. We will describe these shortly. We write static term variables and kind variables in bold for clarity.

The kinding judgement takes the form $\sofkX{\st}{\kappa\moutput}$, where $\kDelta$ and $\kGamma$ are analagous to $\iDelta$ and $\iGamma$ and analagous well-formedness judgements $\kDelta \vdash \kappa$ and $\kDelta \vdash \kGamma$ are also defined. The natural number $n$ is used as a technical device in our semantics to prevent certain forms (those indexed by $n$ in the syntax above) from arising except internally in the semantics (they would have no corresponding concrete syntax); $n$ can be assumed $0$ in all user-defined terms. The dynamic semantics of static terms is defined as a structural operational semantics by the stepping judgement $\sstep{\st}{\argEnv}{\st\moutput}$, the value judgement $\sval{\st}{\argEnv}$ and the error judgement $\serr{\st}{\argEnv}$. Here, $\argEnv$ ranges over \emph{argument environments}, which we will also return to later; we omit the subscript $\argEnv$ in cases where $\argEnv = \cdot; \emptyset; \cdot$. We define the judgement $\smanystep{\st}{\argEnv}{\st\moutput}$ as the reflexive, transitive closure of the stepping judgement and the evaluation judgement $\seval{\st}{\argEnv}{\st'}$ iff $\smanystep{\st}{\argEnv}{\st'}$ and $\sval{\st'}{\argEnv}$. %${\st}{\memD}{\memG}{\aCtx}{\st\moutput}{\memD\moutput}{\memG\moutput}$, where $\memD$, $\memG$ and $\aCtx$ are also technical devices that will be described later; they too can be ignored from the perspective of user code. We write $\st \Downarrow \st'$ iff $\snorm{\st}{\emptyset}{\cdot}{{\rightharpoonup}; \emptyset; \emptyset}{\st'}{\emptyset}{\cdot}$. \emph{Static values} are normalized static terms. Normalization can also raise an error (to indicate a type error in an external term, or a problem in a tycon definition, as we will discuss), indicated by the judgement $\serrX{\st}$. We omit error propagation rules.%We will refer to the relevant rules as we proceed.

\subsection{Types and Tycon Contexts}\label{types}

\begin{figure}[t]
$\begin{array}{lrcllrcl}
\textbf{tycons} & c & ::= & \rightharpoonup ~|~ \tc\\
\textbf{tycon contexts} & \Phi & ::= & \cdot ~|~ \Phi, \tcdef{\tc}{\psi}{\theta}\\
\textbf{tycon structures} & \theta & ::= & \tcstruct{\st}{\omega} & \textbf{tycon sigs} & \psi & ::= & \tcsig{\kappa}{\chi}\\
\textbf{opcon structures} & \omega & ::= & \tcstructn{\st} ~|~ \tcstructc{\omega}{op}{\st}~~~~&\textbf{opcon sigs}& \chi & ::= & \introsig{\kappa} ~|~ \opsigS{\theta}{op}{\kappa}\\
\end{array}$
\caption{Syntax of type constructors. Metavariable $\tc$ ranges over user-defined tycon names.}
\label{syntax-TC}
\end{figure}
